---
title: "Lab 4"
date: "March 2, 2018"
output: 
  html_notebook:
    highlight: tango
    theme: flatly
    toc: yes
    toc_float: yes    
---

RStudio keeps track of the code chunks in the document (see the menu near the bottom left of the code window, to the right of the "line number:column number" counter). Optionally, you can give any chunk a name, for instance "Setup" below. This can make it easier to navigate the code chunks afterwards.

```{r Setup, include = FALSE}

install.packages("statar", dependencies = TRUE)
install.packages('plm', dependencies = TRUE)

library(tidyverse)
library(stargazer)
library(knitr)
library(lubridate)
library(plm)
library(dplyr)
library(statar)

stores <- read.csv('stores.csv')
markets <- read.csv('markets.csv')
  
```


# Data wrangling

## 1 
Let us first inspect the two data sets
```{r Inspect data sets}
stores
markets
```
We want to combine the information in the two data sets. More specifically, we want to add information about the market to each row in the stores data frame. This is achieved by an left join. In order to perform the join we need to extract the year rom the stores data, since we will join using the combination of market and year as key.

```{r Generate year variable}
stores <- mutate(stores, year = year(date))
stores
```


```{r Perform left join}
stores <- left_join(stores, markets, by = c("market", "year"))
stores
```

A `left_join(x,y, by = key)` keeps all observations in "X". A right join would keep all observations in "y", while a full join would keep all observations in "x" and "y". For more information, see chapter 13 in R for Data Science.

## 2 

We want to create a variable with the number of pharmacies in a given market in a given month. First, let us confirm that our data sets contain at most one observation per pharmacy per month

```{r Confirm that there is at most one observation per pharmacy per month}
count(stores, date, phid) %>% summarise(max(n)) 
#Alternatively, we could do 
group_by(stores, date, phid) %>% summarise(n = n()) %>% summarise(max(n))

```

Now, let us count the number of pharmacies per market (in a given month)

```{r Count the number of pharmacies per market}
stores <- group_by(stores, date, market) %>% mutate(ncomp = n()) %>% ungroup()
stores
```

```{r}
summarise(stores, mean = mean(ncomp), min = min(ncomp), max = max(ncomp))
```

## 3

```{r}
stores <- mutate(stores,totpop = f0_19 + f20_66 + f67 +m0_19 + m20_66 + m67) %>%  mutate(popdens = totpop / area)
stores
```


# Present and understand your data

## 4
The "sale" variable is in NOK. More convenient to have it in million NOK

```{r Sale in million NOK}
stores <- mutate(stores, sale = sale / 1000000)
```


```{r Summary statistics}
numvars <- select(stores, storesize, sale, nobs, packages, year, ncomp, totpop, popdens)
stargazer(as.data.frame(numvars), type = 'text')
```

For categorical variables with few values, tabulation is more useful

```{r Tabulate}
count(stores, chain)
```

For categorical variables with many values, such as "phid" tabulation is more difficult. At least it is good to know the number of distinct observations

```{r }
summarise(stores, min_phid = min(phid), max_phid = max(phid), distinct_phid = n_distinct(phid))
```

## 5

When we have many observations, a plain-vanilla scatter is not that informative:

```{r}
ggplot(data = stores) +
  geom_point(mapping = aes(x = nobs, y = sale))
```

Making the points partially transparent helps a bit (but not that much):

```{r}
ggplot(data = stores) +
  geom_point(mapping = aes(x = nobs, y = sale), alpha = .05)
```
A useful variant of a scatter plot is a binscatter (available through the `statar` package). A binscatter plots average values of the y-variable for within bins of the x-variable:

```{r}
ggplot(data = stores, mapping = aes(x = nobs, y = sale)) +
  stat_binmean(n = 50, alpha = 0.5)
```
We could also do something similar "manually":

```{r}
stores <- mutate(stores, nobsgroup = cut(nobs, breaks = seq(0, 9000, by = 1000)))
stores
group_by(stores, nobsgroup) %>% summarise(groupmean = mean(nobs))  %>% 
ggplot() +
  geom_point(mapping = aes(x = nobsgroup, y = groupmean)) +
  scale_x_discrete(breaks=function(x) x[seq(1, length(x), by=2)])
```


Turning now to number of pharmacies in the market:

```{r}
ggplot(data = stores ) +
  geom_point(mapping = aes(x = ncomp, y = sale), alpha = 0.1)
```


```{r}
ggplot(data = stores, mapping = aes(x = ncomp, y = sale)) +
  stat_binmean(n = 25, alpha = 0.5)
```

## 6

```{r}
ggplot(data = stores) + 
  geom_point(mapping = aes(x = packages, y = sale ), alpha = 0.05) + 
  facet_wrap(~ chain)
```


```{r}
ggplot(data = stores, aes(x = packages, y = sale )) + 
  stat_binmean(n = 20, alpha = 0.5) + 
  facet_wrap(~ chain)
```

# Estimation

## 7

As usual we run OLS regressions with the command `lm`:


```{r}
salereg_ncomp <- lm(sale ~ ncomp, data = stores)
stargazer(salereg_ncomp, type = 'text',
          dep.var.labels = "Sales in million NOK",
          covariate.labels = "Number of pharmacies in market")
```

There is a possible relationship between sales and number of competitors in the market. Of course, we should not give this regression a causal interpretation.

## 8

```{r}
stores <- group_by(stores, phid) %>% 
  mutate(demeaned_sale = sale - mean(sale), demeaned_ncomp = ncomp - mean(ncomp)) %>%  
  ungroup()
```

## 9

As before, an ordinary scatter is difficult to read

```{r}
ggplot(data = stores ) +
  geom_point(mapping = aes(x = demeaned_ncomp, y = demeaned_sale), alpha = 0.1)
```

A binscatter:

```{r}
ggplot(data = stores, mapping = aes(x = demeaned_ncomp, y = demeaned_sale)) +
  stat_binmean(n = 200, alpha = 0.5)
```
## 10
```{r}
salereg_ncomp <- lm(demeaned_sale ~ demeaned_ncomp, data = stores)
stargazer(salereg_ncomp, type = 'text',
          dep.var.labels = "Sales in million NOK (demeaned)",
          covariate.labels = "Number of pharmacies in market (demeaned)",
          notes = "Variables are demeaned with respect to pharmacy-level means")
```

## 11
First, we need to create a panel structure for our data set

```{r}
stores_panel = pdata.frame(stores, index=c("phid", "date"))
pdim(stores_panel)
                        
```

We have observations for 433 pharmacies in 94 months. The panel is unbalanced, which means that not all pharmacies are observed in all periods. (Why could that be?) Fixed effects ("whithin") estimatoin is done with the `plm` package.


```{r}
salereg_ncomp_fe <- plm(sale ~ ncomp, data = stores_panel, model = 'within')
stargazer(salereg_ncomp_fe, type = 'text',
          dep.var.labels = "Sales in million NOK",
          covariate.labels = "Number of pharmacies in market"
          )
```
We see that the results are the same as in task 10 (which is just a manual way of implementing the fixed effects regression). The result is very different from the simple OLS in task 7: the estimated coefficient has the opposite sign.

## 12

A regression with market-level fixed effects can be implemented bt adding dummies for the markets (and then running an OLS-regression).

```{r}
salereg_ncomp_market_fe <- lm(sale ~ ncomp + factor(market), data = stores)
stargazer(salereg_ncomp_market_fe, type = 'text',
          dep.var.labels = "Sales in million NOK",
          covariate.labels = "Number of pharmacies in market",
          keep = "ncomp",
          notes = "Dummy variables for markets included in the regression but excluded from table"
          )
```

The negative effect of more competitors on sales is now estimated to be even stronger. 

# Extra

## a

```{r Get average number of pharmacies per yer and market}
market_years <- group_by(stores, year, market) %>%  
  summarise(ncomp = mean(ncomp),
            totpop = mean(totpop))
```


```{r Do a left-join with the original markets data to get variables that are constant across years}
market_years <- left_join(market_years, markets, by = c("year","market")) 
market_years
```

(Could also have summarized the population variables, but proposed method saves some typing.)

## b
```{r}
ggplot(market_years, mapping = aes(x = totpop, y = ncomp)) + 
  geom_point() +
  geom_smooth(formula = y ~ x, method='lm') + 
  geom_smooth(formula = y ~ poly(x, 2), method = 'lm')
```
Market-years with large markets (high population) tend to hanve many competitors. The relationship seems to be concave. We expect a concave relionshiop between number of competitors and market size if more competitors leads to more intense competition (cf. the with the discussion in lecture 7, slide 29-30.)


## c

```{r Demeaning the variables}
market_years <- group_by(market_years, market) %>%
  mutate(demeaned_ncomp = ncomp - mean(ncomp), demeaned_totpop = totpop - mean(totpop)) %>% 
  ungroup()
```



```{r Inspect the data frame}
market_years
```

```{r Plot the relationship between demeaned number of pharmacies and demeaned market size}
ggplot(market_years, mapping = aes(x = demeaned_totpop, y = demeaned_ncomp)) + 
  geom_point() +
  geom_smooth(formula = y ~ x, method='lm') + 
  geom_smooth(formula = y ~ poly(x, 2), method = 'lm')
```
The relationship is similar as the one found in b). Within markets, there tends to be more competitors in periods where the market size is large (measured by population). This makes intuitive sense. Note however that the relationship is close to being linear.

## d


```{r}
ncomp_dem <- lm(ncomp ~  f0_19 + f20_66 + f67 + m0_19 + m20_66 + m67, data = market_years)
stargazer(ncomp_dem , type = 'text')
```

Not a very clear picture.

## e
A fixed effects estimation will indicate whether the y-variable is larger/smaller than the average (for the market) in periods where the x-variables are larger/smaller than average (for the market). It is therefore well suited for this task.


```{r Declear the data to be a paneø}
market_panel = pdata.frame(market_years, index=c("market", "year"))
pdim(market_panel)
```

We have an unbalanced panel with 126 markets and 8 periods (years).

```{r Fixed effects total population}
ncomp_pop_fe <- plm(ncomp ~  totpop, data = market_panel, model = 'within')
stargazer(ncomp_pop_fe, type = 'text')
```

Increases in population seem to be assosiated with entry.

```{r Fixed effects demographics}
ncomp_dem_fe <- plm(ncomp ~  f0_19 + f20_66 + f67 + m0_19 + m20_66 + m67, data = market_panel, model = 'within')
stargazer(ncomp_dem_fe, type = 'text')
```
More senior citizens are associated with entry. For other age groups, picture is not so clear.

## f