---
title: "Lab 4"
date: "March 2, 2018"
output: 
  html_notebook:
    highlight: tango
    theme: flatly
    toc: yes
    toc_float: yes    
---

RStudio keeps track of the code chunks in the document (see the menu near the bottom left of the code window, to the right of the "line number:column number" counter). Optionally, you can give any chunk a name, for instance "Setup" below. This can make it easier to navigate the code chunks afterwards.

```{r Setup, include = FALSE}

#Uncomment the next line if the `statar` package is not installed
#install.packages("statar", dependencies = TRUE)

library(tidyverse)
library(stargazer)
library(knitr)
library(lubridate)
library(plm)
library(dplyr)
library(statar)

stores <- read.csv('stores.csv')
markets <- read.csv('markets.csv')
  
```


# Data wrangling

## 1 
Let us first inspect the two data sets
```{r Inspect data sets}
stores
markets
```
We want to combine the information in the two data sets. More specifically, we want to add information about the market to each row in the stores data frame. This is achieved by an left join. In order to perform the join we need to extract the year rom the stores data, since we will join using the combination of market and year as key.

```{r Generate year variable}
stores <- mutate(stores, year = year(date))
```


```{r Perform left join}
#stores <- left_join([to do])
```

A `left_join(x,y, by = key)` keeps all observations in "X". A right join would keep all observations in "y", while a full join would keep all observations in "x" and "y". For more information, see chapter 13 in R for Data Science.

## 2 

We want to create a variable with the number of pharmacies in a given market in a given month. First, let us confirm that our data sets contain at most one observation per pharmacy per month

```{r Confirm that there is at most one observation per pharmacy per market}
count(stores, date, phid) %>% summarise(max(n)) 
```

Now, let us count the number of pharmacies per market (in a given month)

```{r Count the numer of pharmacies per market}
#stores <- [to do]

```

## 3

To do

# Present and understand your data

## 4
The "sale" variable is in NOK. More convenient to have it in million NOK

```{r Sale in million NOK}
stores <- mutate(stores, sale = sale / 100000)
```

Summary statistics: A good option is to use the *stargazer* package

## 5

When we have many observations, a plain-vanilla scatter is not that informative:

```{r}
ggplot(data = stores) +
  geom_point(mapping = aes(x = nobs, y = sale))
```

Making the points partially transparent helps a bit (but not that much):

```{r}
ggplot(data = stores) +
  geom_point(mapping = aes(x = nobs, y = sale), alpha = .05)
```
A useful variant of a scatter plot is a binscatter (available through the *statar* package). A binscatter plots average values of the y-variable for within bins of the x-variable:

```{r}
ggplot(data = stores, mapping = aes(x = nobs, y = sale)) +
  stat_binmean(n = 50, alpha = 0.5)
```

Relationship between sales and number of competitors: To do

## 6

Hint: use `facet_wrap`

# Estimation

## 7

As usual we run OLS regressions with the command `lm`:

## 8
To do

## 9
To do

## 10
To do

## 11
First, we need to create a panel structure for our data set

```{r}
stores_panel = pdata.frame(stores, index=c("phid", "date"))
pdim(stores_panel)
                        
```

We have observations for 433 pharmacies in 94 months. The panel is unbalanced, which means that not all pharmacies are observed in all periods. (Why could that be?) Fixed effects ("whithin") estimatoin is done with the *plm* package.


## 12

# Extra

## a
## b
## c
## d
## e
## f